A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-05-15 papers/water-SAFT/paper.tex
Added density_calc.py and data file.  Made minor changes to paper. Fixed error due to assert in src/lattice.cpp by declaring vol0. Tested everything on quipu.

A large fraction of interesting chemistry---including all of molecular
biology---takes place in aqueous solution.  However, while quantum
chemistry now enables us to calculate the ground state energies of
large molecules in vacuum, prediction of the free energy of even the
smallest molecules in the presence of a solvent poses a continuing
challenge due to the complex structure of a liquid and the
computational cost of \emph{ab initio} molecular
dynamics~\cite{car1985, grossman2004}.  The current state-of-the art
in \emph{ab initio} molecular dynamics is limited to around 64 water
molecules per unit cell, which places the minimum solute concentration
at 0.84 molar and allows contact between the second solvation shell of
neighboring solutes~\cite{izvekov2005, choe2007}.  On top of this,
standard \emph{ab initio} methods using classical molecular dynamics
without van der Waals corrections strongly over-structure water, to
the point that ice melts at over 400K\cite{yoo2009phase}!  There has
been a recent flurry of publications implicating van der Waals effects
as significant in reducing this
over-structuring\cite{lin2009importance, wang2011density,
  mogelhoj2011ab, jonchiere2011van}.  It has also been bound that
inclusion of nuclear quantum effects can have a similar
effect\cite{morrone2008nuclear}.  Each of these corrections imposes an
additional computational burden on an approach that is already
feasible only for very small number of water molecules. A more
efficient approach is needed in order to study nanoscale and
hydrophobic solutes.

@@

Due to its large dielectric constant, small molecular size and strong
hydrogen-bonding network, water as ``the universal solvent'' is as
interesting as its anomalous thermodynamic properties.

@@

\section{Theory and Methods}
We construct a new classical density functional for water.  We
start with a treatment of each individual water molecule as a purely
classical hard sphere and add association terms based on Statistical 
Associating Fluid Theory (SAFT). Water is unlike simple
fluids (which are mainly defined by van der Waals forces and weak electrostatic
interations) because it requires direct treatment of association effects. The strong 
hydrogen bonds and polar
interactions in water are ideal for describing with SAFT, which was designed for
use with hydrogen-bonding fluids. It is based on Wertheim's first-order
thermodynamic perturbation theory (TPT1)
\cite{wertheim1984fluidsI,wertheim1984fluidsII,wertheim1986fluidsIII,
wertheim1986fluidsIV} which provides a useful way to account for strong
associative interations between molecules. SAFT can account for the association energies
as well as any effects due to clusters or chains forming \fixme{not true}
in the liquid\cite{muller2001molecular}. SAFT assumes a repulsive hard
sphere core for each atom with a number of short-ranged off center
attractive sites. This model can be applied to give an equation of state for
various systems even inhomogeneous fluids.
\fixme{Expand this a
  bit to accurately describe SAFT and its advantages.}

>>>>

A large fraction of interesting chemistry---including all of molecular
biology---takes place in aqueous solution.  However, while quantum
chemistry now enables us to calculate the ground state energies of
large molecules in vacuum, prediction of the free energy of even the
smallest molecules in the presence of a solvent poses a continuing
challenge due to the complex structure of a liquid and the
computational cost of \emph{ab initio} molecular
dynamics~\cite{car1985, grossman2004}.  The current state-of-the art
in \emph{ab initio} molecular dynamics is limited to around 64 water
molecules per unit cell, which places the minimum solute concentration
at 0.84 molar and allows contact between the second solvation shell of
neighboring solutes~\cite{izvekov2005, choe2007}.  On top of this,
standard \emph{ab initio} methods using classical molecular dynamics
without van der Waals corrections strongly over-structure water, to
the point that ice melts at over 400K\cite{yoo2009phase}!  There has
been a recent flurry of publications implicating van der Waals effects
as significant in reducing this
over-structuring\cite{lin2009importance, wang2011density,
  mogelhoj2011ab, jonchiere2011van}.  It has also been found that
inclusion of nuclear quantum effects can provide similar improvements
\cite{morrone2008nuclear}.  Each of these corrections imposes an
additional computational burden on an approach that is already
feasible only for a very small number of water molecules. A more
efficient approach is needed in order to study nanoscale and
hydrophobic solutes.

@@

Due to its large dielectric constant, small molecular size and strong
hydrogen-bonding network, water as ``the universal solvent'' is as
interesting as its anomalous thermodynamic properties.  \fixme{what
  the heck?}

@@

\section{Theory and Methods}
We construct a new classical density functional for water by treating
each individual water molecule as a purely
classical hard sphere and adding association terms based on SAFT. Water
is unlike simple fluids (which are mainly defined by van der Waals
forces and weak electrostatic interations) because it requires direct
treatment of association effects. The strong 
hydrogen bonds and polar
interactions in water are ideal for describing with SAFT, which was designed for
use with hydrogen-bonding fluids.
